Quantitative Auditing of AI Fairness with Differentially Private Synthetic Data

### 2025-01-15 paper/paper.tex
Major

For instance, investigative journalists at ProPublica uncovered bias in the COMPAS system, which is used in the United States criminal justice system to predict recidivism. Their analysis revealed that the COMPAS system disproportionately affects certain demographic groups.

@@

One effective approach to achieving fairness in AI systems is to conduct audits. Quantitative fairness auditing involves evaluating these systems using various fairness metrics to identify potential biases and inequalities. Some commonly proposed metrics for assessing AI system fairness include demographic parity, equalized odds, and accuracy equality. These metrics provide a quantitative foundation for addressing fairness concerns in AI models.

A well-documented example of fairness auditing is the investigation of the COMPAS system. Journalists from ProPublica revealed that COMPAS exhibited bias against certain demographic groups, particularly African Americans. The system was more likely to falsely classify African American defendants as high risk for reoffending compared to their white counterparts. This finding not only underscored the necessity of fairness auditing but also highlighted the potential harm of deploying biased AI systems in critical areas.

@@

While fairness auditing is essential, it presents significant challenges, particularly when it involves sensitive data. To conduct audits, auditors may rely on access to real datasets containing sensitive information, raising significant security and privacy concerns. Auditors often require access to real-world datasets containing sensitive attributes, such as race, gender, or income, to evaluate an AI system's fairness. However, this reliance on sensitive data raises concerns about data privacy and security.

@@

Moreover, analyses on real data runs the risk of data inference attacks, where confidential information may be deduced from seemingly harmless or aggregated data. Even if the real data is not directly exposed, the results of fairness analyses can sometimes reveal sensitive information through inference attacks. For example, attackers could combine publicly available audit results with external datasets to infer private details about individuals in the original dataset.

@@

Various technologies have been developed to mitigate these risks, such as differential privacy. Differential privacy is a mathematical framework that ensures the inclusion or exclusion of any individual in a dataset has a minimal and quantifiable impact on the overall results of data analysis. This is achieved by introducing carefully calibrated random noise into the computations derived from the data, ensuring that individual records remain indistinguishable while preserving aggregate patterns and trends.

The concept of differential privacy is particularly valuable in scenarios where sensitive information, such as health records or financial data, is involved. For example, a differential privacy mechanism could allow researchers to analyze patterns in patient health outcomes without exposing any single patient's medical history. This approach strikes a balance between data utility and privacy, enabling meaningful analysis while safeguarding individuals against privacy breaches or inference attacks.

@@

Among the generation techniques, differentially private synthetic data offers a way to investigate the data properties under differential privacy guarantees. It stands out for it strong differential privacy guarantees when generating synthetic data. This method combines data synthesis with the principles of differential privacy, ensuring that no single individual's information significantly influences the synthetic dataset. By introducing controlled randomness during generation, it provides strong privacy guarantees while preserving the overall utility of the data.

Differentially private synthetic data is particularly valuable in sensitive domains like healthcare and finance, enabling researchers to analyze patterns or train AI models without exposing personal details. It also supports fairness auditing by allowing the examination of datasets and models without revealing sensitive attributes. This dual focus on privacy and utility makes it a valuable tool.

\subsection{Differentially Private Synthetic Data for Fairness Auditing}

@@

@@

% contribution

In this paper, we propose a novel framework for fairness auditing that leverages differentially private synthetic data. Our framework enables auditors to evaluate the fairness of AI systems without exposing sensitive information, addressing both privacy and security concerns.

Our work examines the capabilities of existing differentially private synthetic data generation technologies in preserving fairness properties. Through empirical experiments on multiple real datasets, we aim to determine the effectiveness of differentially private synthetic data in this respect.

>>>>

For instance, investigative journalists at ProPublica uncovered bias in the COMPAS system, which is used in the United States criminal justice system to predict recidivism. Their analysis revealed that the COMPAS system disproportionately discriminated against certain demographic groups.

@@

One effective approach to achieving fairness in AI systems is to conduct audits. Quantitative fairness auditing involves evaluating these systems using various fairness metrics to identify potential biases and inequalities. Some commonly proposed metrics for assessing AI system fairness include demographic parity, equalized odds, and accuracy equality. These metrics provide a quantitative basis for addressing fairness concerns in AI models.

A well-documented example of fairness auditing is the investigation of the COMPAS system. Journalists from ProPublica revealed that COMPAS exhibited bias against certain demographic groups, particularly African Americans. The system was more likely to falsely classify African American defendants as high risk compared to their Caucasian counterparts. This finding not only underscored the necessity of fairness auditing but also highlighted the potential harm of deploying biased AI systems in critical areas.

@@

While fairness auditing is essential, it presents significant challenges, particularly when it involves sensitive data. To conduct audits, auditors may rely on access to real datasets containing sensitive information. Auditors often require access to real-world datasets containing sensitive attributes, such as race, gender, or income, to evaluate an AI system's fairness. However, this reliance on sensitive data raises concerns about data security and privacy.

@@

Moreover, analyses of real data run the risk of data inference attacks, where confidential information may be deduced from seemingly harmless or aggregated data. Even if the real data is not directly exposed, the results of analyses can sometimes reveal sensitive information through inference attacks. For example, attackers could combine publicly available audit results with external datasets to infer private details about individuals in the original dataset.

@@

Various technologies have been developed to mitigate privacy risks, such as differential privacy. Differential privacy is a mathematical framework that ensures the inclusion or exclusion of any individual in a dataset has a minimal and quantifiable impact on the overall results of data analysis. This is achieved by introducing carefully calibrated random noise into the computations derived from the data, ensuring that individual records remain indistinguishable while preserving aggregate patterns and trends.

The concept of differential privacy is particularly valuable in scenarios where sensitive information, such as health records or financial data, is involved. For example, a differential privacy mechanism could allow researchers to analyze patterns in patient health data without exposing any single patient's medical history. This approach strikes a balance between data utility and privacy, enabling meaningful analysis while safeguarding individuals against privacy breaches or inference attacks.

@@

Among the generation techniques, differentially private synthetic data offers a way to investigate the data properties under differential privacy guarantees. It stands out for it strong differential privacy guarantees when generating synthetic data. This method combines data synthesis with the principles of differential privacy, ensuring that individuals' sensitive information cannot be inferred from the synthetic data. By introducing controlled randomness during generation, it provides strong privacy guarantees while preserving the overall utility of the data.

Differentially private synthetic data is particularly valuable in sensitive domains like healthcare and finance, enabling researchers to analyze patterns or identify trends without exposing personal details. It also supports fairness auditing by allowing the examination of datasets and models without revealing sensitive attributes. This dual focus on privacy and utility makes it a valuable tool.

@@

We present an innovative solution to the issue of sensitive data access: auditors generate synthetic data immediately upon receiving the real dataset, which they retain for auditing purposes while discarding the original data.

@@

Our work examines the capabilities of existing differentially private synthetic data generation technologies in preserving fairness properties. Through empirical experiments on multiple real datasets, we aim to determine the effectiveness of differentially private synthetic data in fairness auditing.

We perform a comparative analysis of fairness metrics computed on real and synthetic data across multiple real-world datasets, including Adult, COMPAS, and Diabetes. By examining the consistency and differences between these metrics, we evaluate how well synthetic data captures the fairness characteristics of real data.
